\documentclass{article}
\usepackage{graphicx} %for images
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{animate}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{animate}


\usepackage{minted}
\definecolor{codebg}{RGB}{243, 243, 243}

\newminted{python} {
    style=tango,
    bgcolor=codebg,
    linenos,
    fontsize=\small,
    breaklines
}
\setlength{\parindent}{0pt}

\renewcommand{\refname}{}


\title{Modeling Travel in Gravitational Fields}
\author{Brian Zhou, Minh Nguyen, James Overcash} 
\date{April 2026}

\begin{document}

\maketitle

\section{Conventions}
Bolded variables are vectors. The multiplication of two scalars $a$ and $b$ is always represented implicitly as $ab$ while the scaling of a vector $\bm{v}$ by a scalar $k$ is always represented explicitly as $k \cdot \bm{v}$. Newton's notation is used for all derivatives with respect to time, and Leibniz's notation is used for all derivatives with respect to other quantities.

\section{Introduction}
We consider a rocket with a single rear-mounted thruster traveling through space. The rocket body is modeled as a uniformly dense rod of negligible thickness with length $L$ meters and mass $m_b$ kilograms, and it carries an additional $m_f$ kilograms of fuel. The thruster consumes fuel at a constant rate of $s_f$ kilograms per second and, while fuel remains, produces a constant thrust force of $T$ Newtons in the direction opposite to the rocket's vector-orientation.\\

The rocket moves in a two-dimensional planar space containing $N$ uniformly dense circular planets. The planets have fixed positions $\bm{p}_1,\bm{p}_2, \dots, \bm{p}_N$, masses $m_1, m_2, \dots, m_N$, and radii $r_1, r_2, \dots, r_N$. The mass of each planet exerts a gravitational force affecting the rocket's motion.\\

Given the rocket's initial position $\bm{p}_0$, velocity $\bm{v}_0$, and angle-orientation $\theta_0$ (counter-clockwise from the $x$-axis), we model its trajectory under the combined effects of planetary gravity and thrust using the Euler-Cromer method.

\section{Analysis}
Let $\bm{p}(t), \bm{v}(t), \theta(t), $ and $m(t)$ denote the position, velocity, angle-orientation, and mass, respectively, of the rocket at time $t$. Additionally, define $\bm{u}(t) = (\cos\theta(t), \sin\theta(t))$ to describe the vector-orientation of the rocket at time $t$. We will model the rocket's path using classical Newtonian mechanics. Thus, the linear acceleration $\ddot{\bm{p}}(t)$ and linear velocity $\dot{\bm{p}}(t)$ can be related to the net force exerted on the rocket $\bm{F}_\text{net}(t)$ by Newton's second law 
\[
\bm{F}_{\text{net}}(t) = m(t) \cdot \ddot{\bm{p}}(t) + \dot{m}(t) \cdot \dot{\bm{p}}(t). \qquad\cite{newtons-laws-of-motion}
\]
Since the thruster consumes fuel at a constant rate as long as there is fuel, there exists a time $t_e$ where the rocket no longer experiences a thrust. Using algebra,
\begin{align*}
t_e &= \frac{m_f}{s_f}\\
m(t) &= \begin{cases}
    m_b + m_f - s_ft & \text{if } t < t_e\\
    m_b & \text{otherwise}
\end{cases}\\
\dot{m}(t) &= \begin{cases}
    -s_f & \text{if } t < t_e\\
    0 & \text{otherwise}
\end{cases}
\end{align*}
Similarly, if we let $\bm{F}_T(t)$ describe the thrust at time $t$, it follows that
\[
\bm{F}_T(t) = \begin{cases}
    -T \cdot \bm{u}(t) & \text{if } t < t_e\\
    0 & \text{otherwise}
\end{cases}
\]
If we let $\bm{F}_G(t)$ describe the net gravitational force applied to the rocket at time $t$ and $\bm{F}_{G_i}(t)$ describe the specific gravitational force applied to the rocket due to the $i$th planet at time $t$, then it follows that
\[
\bm{F}_G(t) = \sum_{i=1}^{N}\bm{F}_{G_i}(t).
\]
Using Newton's law of universal gravitation \cite{gehrels}, we can expand these definitions
\begin{align*}
\bm{F}_{G_i}(t) &= \frac{Gm(t)m_i}{\lvert \bm{p}(t) - \bm{p}_i \rvert^3} \cdot (\bm{p}(t) - \bm{p}_i)\\
\therefore \bm{F}_G(t) &= Gm(t)\sum_{i=1}^{N}\frac{m_i}{\lvert \bm{p}(t) - \bm{p}_i \rvert^3} \cdot (\bm{p}(t) - \bm{p}_i)
\end{align*}
where $G \approx 6.674 \times 10^{-11}$ is Newton's gravitational constant. Since gravity and thrust are the only forces acting upon the rocket,
\[
\bm{F}_\text{net}(t) = \bm{F}_T(t) + \bm{F}_G(t)
\]
which we can use to isolate the linear acceleration $\ddot{\bm{p}}(t)$
\begin{align*}
m(t) \cdot \ddot{\bm{p}}(t) + \dot{m}(t) \cdot \dot{\bm{p}}(t) &= \bm{F}_T(t) + \bm{F}_G(t) \\
\ddot{\bm{p}}(t) &= \frac{1}{m(t)} \cdot \bm{F}_T(t) + \frac{1}{m(t)} \cdot \bm{F}_G(t) - \frac{\dot{m}(t)}{m(t)} \cdot \dot{\bm{p}}(t).
\end{align*}
% Converting the second-order differential equation into a system of first-order ones, we have

% \begin{align*}
% \dot{\bm{x}} &= \dot{\bm{p}} \\
% \dot{\bm{v}} &= \ddot{\bm{p}} = \frac{\bm{F}_T(t)}{m(t)} + G\sum_{i=1}^{N}\frac{m_i}{\lvert \bm{p}(t) - \bm{p}_i \rvert^3} \cdot (\bm{p}(t) - \bm{p}_i) - \frac{\dot{m}(t)}{m(t)} \cdot \dot{\bm{p}}(t)
% \end{align*}

The Euler-Cromer method can be applied to systems of differential equations of the form
\begin{align*}
    \bm{\dot{x}} &= f(t, v) \\
    \bm{\dot{v}} &= g(t, x)
\end{align*}
where $\bm{x}$ and $\bm{v}$ respectively represent position and velocity, making it applicable for our model. The Euler-Cromer method approximates a discrete solution by iterating using the following equations
\begin{align*}
    \bm{v}_{n+1} &= \bm{v}_n + \bm{a}(x_n, t_n)\Delta t \\
    \bm{x}_{n+1} &= \bm{x}_{n} + \bm{v}_{n+1}\Delta t
\end{align*}
where $\Delta t$ is the time step; $v_n, x_n$ represent velocity and position after $n$ time steps; and $\bm{a}$ represents the acceleration function. \\

The difference between the Euler-Cromer method and the standard Euler method is that it uses $v_{n+1}$ to solve for $x_{n+1}$ instead of $v_n$. This difference means energy nearly completely conserved within the system when applying this method, whereas energy tends to continually increase when applying the standard Euler method.\cite{euler} So, the Euler-Cromer method is more accurate than the standard Euler method for this application. \\

Thus, we will be finding an approximate solution using
\begin{align*}
    \dot{\bm{p}}_{n+1} &= \dot{\bm{p}}_n + \left(
    \frac{\bm{F}_T(t)}{m(t)} + G\sum_{i=1}^{N}\frac{m_i}{\lvert \bm{p}_n - \bm{p}_i \rvert^3} \cdot (\bm{p}_n - \bm{p}_i) - \frac{\dot{m}(t)}{m(t)} \cdot \dot{\bm{p}}_n \right)\Delta t \\
    \bm{p}_{n+1} &= \bm{p}_{n} + \dot{\bm{p}}_{n+1}\Delta t.
\end{align*}

\vspace{1cm}


Furthermore, since the rocket is not a point mass, it experiences a torque about its center of mass due to the gravitational forces acting upon it. Since the rocket and planets exist in a two-dimensional planar space, all rotation occurs about a single axis perpendicular to the plane of the space. Therefore, the rotational acceleration $\ddot{\theta}(t)$ can be related to the net torque exerted on the rocket $\tau_\text{net}(t)$ by Newton's second law for rotation 
\[
\tau_\text{net}(t) = I(t)\ddot{\theta}(t) \qquad\cite{openstax-newtons-second-law-of-rotation}
\]
where $I(t)$ denotes the moment of inertia of the rocket at time $t$. Modelling the rocket as a uniform rod with negligible thickness, its moment of inertia around the center is.
\[
I(t)=\frac{1}{12}m(t)L^2. \qquad\cite{hyperphysics-rod-inertia-formula}
\]
(We will continue with the angular portion of the simulation in the final draft.)

\section{Reality Check}
Python code in the appendices will model this system for the reality check.
% \begin{pythoncode}
% r = Rocket(Vector2d(0, 0), Vector2d(sqrt(10 * G), 0), 100)
% planets = [
%     Planet(Vector2d(10, 10), 100, 2),
%     Planet(Vector2d(25, 60), 100, 2),
%     Planet(Vector2d(50, 90), 100, 2),
%     Planet(Vector2d(100, 100), 100, 2),
% ]
% \end{pythoncode}
The following animation visualizes the trajectory of the rocket. The black circle represents a point-mass rocket, the light gray circles represent the planets, and the multicolor curve represents the trajectory of the rocket. The animation will only play when opening this file using Adobe Acrobat or PDF viewers with similar capabilities. The link below the image links to a higher-quality version of the same animation which does not require specific viewing applications to play.\\

\textbf{Figure 1}\\
\textit{Reality check on a four-planet system}
\begin{center}
    % \animategraphics[loop,autoplay,width=0.5\textwidth]{12}{test/image-}{1}{121}
    % \url{https://github.com/Brian21347/rocket-modeling/blob/main/animations/animation.gif}
\end{center}

\section{Conclusion}
We expanded upon Euler's method of approximating solutions to differential equations learned in the course to model the trajectory of an ideal rocket in a system of four planets via the Euler-Cromer method. As the Euler-Cromer method is first-order accurate, one limitation is that the time steps must be incredibly small for high accuracy simulations and will require more computational power than the Runge-Kutta method to achieve the same accuracy. One possible extension of the project would be to consider the rocket and planets as inhabiting a three-dimensional space instead of a two-dimensional one, where rotation would naturally get more complex and will involve the tensor definition of the moment of inertia.

\section{References}
Results in Figure 1 were obtained using a Python script provided in Appendix 1.

\vspace{-1cm}

\begin{thebibliography}{99}

\bibitem{gehrels}
Gehrels, N. (n.d.). Newton’s law of gravitation derivation. NASA. \url{https://imagine.gsfc.nasa.gov/observatories/learning/swift/classroom/law_grav_derivation.html}
\bibitem{newtons-laws-of-motion}
Newton’s laws of motion. (2026). In Wikipedia. \url{https://en.wikipedia.org/w/index.php?title=Newton%27s_laws_of_motion&oldid=1336713731}
\bibitem{euler}
Semi-implicit Euler method. (2026). In Wikipedia. \url{https://en.wikipedia.org/w/index.php?title=Semi-implicit_Euler_method&oldid=1333480764}
\bibitem{openstax-newtons-second-law-of-rotation}
Moebs, W., Ling, S. J., \& Sanny, J. \textit{Newton’s Second Law for Rotation}. University Physics Volume 1, OpenStax. 
\url{https://openstax.org/books/university-physics-volume-1/pages/10-7-newtons-second-law-for-rotation}
\bibitem{hyperphysics-rod-inertia-formula}
Nave, R. \textit{Moments of Inertia for Common Shapes}. HyperPhysics, Georgia State University. 
\url{http://hyperphysics.phy-astr.gsu.edu/hbase/mi2.html}
\end{thebibliography}



\section{Appendices}
\subsection{model.py}
This file contains the code modeling the trajectory of the rocket.
% \begin{lstlisting}[style=python]
% \begin{pythoncode}
% from interfaces import *
% import tqdm


% class Model:
%     def __init__(
%         self,
%         rocket: Rocket,
%         planets: Sequence[Planet],
%         simulation_hours: int,
%         dt: int,
%     ) -> None:
%         self.rocket = rocket
%         self.planets = planets
%         self.simulation_seconds = simulation_hours
%         self.dt = dt
%         self.path = []

%     def crash_checking(self) -> Planet | None:
%         for planet in self.planets:
%             if (
%                 abs(self.rocket.position.x - planet.position.x)
%                 + abs(self.rocket.position.y - planet.position.y)
%                 > planet.radius
%             ):
%                 continue
%             dist = self.rocket.position.dist(planet.position)
%             if dist <= planet.radius:
%                 return planet
%         return None

%     def estimate_path(self) -> Planet | None:
%         dmassfuel = -self.rocket.thrust.mag() / self.rocket.speed_fuel * self.dt
%         self.path.append(self.rocket.position)
%         for _ in tqdm.tqdm(range(self.simulation_seconds // self.dt)):
%             if self.rocket.mass_fuel == 0:
%                 thrust = Vector2d(0, 0)
%             elif self.rocket.mass_fuel + dmassfuel < 0:
%                 thrust_percent = self.rocket.mass_fuel / -dmassfuel
%                 thrust = self.rocket.thrust * thrust_percent
%                 self.rocket.mass_fuel = 0
%             else:
%                 thrust = self.rocket.thrust
%                 self.rocket.mass_fuel += dmassfuel

%             rocket_mass = self.rocket.mass_fuel + self.rocket.mass_ship
%             self.rocket.velocity += (
%                 thrust / rocket_mass
%                 + self.calc_force()
%                 - dmassfuel / rocket_mass * self.rocket.velocity
%             ) * self.dt
%             self.rocket.position += self.rocket.velocity * self.dt
%             self.path.append(self.rocket.position)
%             if (crash_dest := self.crash_checking()) != None:
%                 return crash_dest

%     def calc_force(self) -> Vector2d:
%         force = Vector2d([0, 0])
%         for planet in self.planets:
%             dist = self.rocket.position.dist(planet.position)
%             force += (planet.position - self.rocket.position) * planet.mass / dist**3
%         return G * force

% % \end{pythoncode}

% \subsection{interfaces.py}
% This file contains the interfaces used in model.py.
% \begin{pythoncode}
% from dataclasses import dataclass
% from typing import overload
% from collections.abc import Sequence, Callable
% from math import sqrt


% @dataclass
% class Rocket:
%     position: Vector2d
%     mass_ship: float
%     velocity: Vector2d

%     mass_fuel: float = 0
%     speed_fuel: float = 0
%     thrust: Vector2d = Vector2d(0, 0)


% @dataclass
% class Planet:
%     position: Vector2d
%     mass: float
%     radius: float = 0


% class Vector2d:
%     pos: tuple[float, float]

%     @overload
%     def __init__(self, x: float, y: float, /) -> None: ...

%     @overload
%     def __init__(self, pos: Sequence[float,], /) -> None: ...

%     def __init__(self, *args) -> None:
%         if len(args) > 2:
%             raise ValueError()
%         if len(args) == 1:
%             self.pos = tuple(args[0])
%         elif len(args) == 2:
%             self.pos = args

%     def dist(self, other: 'Vector2d'):
%         return sqrt((self.x - other.x) ** 2 + (self.y - other.y) ** 2)
%     def mag(self):
%         return sqrt(self.x ** 2 + self.y ** 2)
%     @property
%     def x(self):
%         return self.pos[0]
%     @x.setter
%     def x(self, val: float):
%         self.pos = val, self.pos[1]
%     @property
%     def y(self):
%         return self.pos[1]
%     @y.setter
%     def y(self, val: float):
%         self.pos = self.pos[0], val
%     def __str__(self) -> str:
%         return f'{self.x} {self.y}'
%     def __add__(self, other: 'Vector2d'):
%         return Vector2d(self.x + other.x, self.y + other.y)
%     def __sub__(self, other: 'Vector2d'):
%         return Vector2d(self.x - other.x, self.y - other.y)
%     def __mul__(self, other: float):
%         return Vector2d(self.x * other, self.y * other)
%     def __rmul__(self, other: float):
%         return self * other
%     def __truediv__(self, other: float):
%         return Vector2d(self.x / other, self.y / other)
% \end{pythoncode}

\iffalse
Next, we calculate the thrust force $\bm{F}_T$ of the rocket. We assume that the rocket ejects fuel at a constant rate $\frac{dm}{dt}$ with a constant exhaust speed $\lvert \bm{v}_e \rvert$. Therefore, according to the Tsiolkovsky rocket equation, $\bm{F}_T$ is of the form
\begin{align*}
    \bm{F}_T(t) &= \frac{dm}{dt} \cdot \bm{v}_e\\
\end{align*}
\fi
\iffalse
    This suggests the thrust force of the rocket expressed as a function of time is:
    \[
    \bm{F}_T(t) = \begin{cases}
        \bm{T} & \text{if } t \leq t_b \\
        \bm{0} & \text{if } t > t_b
    \end{cases}
    \]
    
    Then, the equation is
    \[
    F(t)= \begin{cases}
    GM\sum\limits_{i=1}^{N}\frac{m_i}{\lvert \bm{p}(t) - \bm{p}_i \rvert^3} \cdot (\bm{p}(t) - \bm{p}_i) + \bm{T} & \text{if } t \leq t_b \\
    GM\sum\limits_{i=1}^{N}\frac{m_i}{\lvert \bm{p}(t) - \bm{p}_i \rvert^3} \cdot (\bm{p}(t) - \bm{p}_i) & \text{if } t > t_b
    \end{cases}
    \]
\fi
\end{document}

